---
output:
  rmarkdown::github_document
bibliography: "README-refs.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```


# R/`medshift`

[![Travis-CI Build Status](https://travis-ci.org/nhejazi/medshift.svg?branch=master)](https://travis-ci.org/nhejazi/medshift)
[![AppVeyor Build  Status](https://ci.appveyor.com/api/projects/status/github/nhejazi/medshift?branch=master&svg=true)](https://ci.appveyor.com/project/nhejazi/medshift)
[![Coverage Status](https://img.shields.io/codecov/c/github/nhejazi/medshift/master.svg)](https://codecov.io/github/nhejazi/medshift?branch=master)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)

> Targeted Learning of the Causal Effects of Stochastic Interventions

__Authors:__ [Nima Hejazi](https://nimahejazi.org) and [Ivan
Diaz]()

---

## What's `medshift`?

The `medshift` R package is designed to provide facilities to...

---

## Installation

Install the most recent _stable release_ from GitHub via
[`devtools`](https://www.rstudio.com/products/rpackages/devtools/):

```{r gh-master-installation, eval = FALSE}
devtools::install_github("nhejazi/medshift", build_vignettes = FALSE)
```

---

## Example

To illustrate how `medshift` may be used to ascertain the effect of a treatment,
consider the following example:

```{r example, message=FALSE, warning=FALSE, eval=FALSE}
library(medshift)
library(condensier)
set.seed(429153)

# simulate simple data for tmle-shift sketch
n_obs <- 1000  # number of observations
n_w <- 1  # number of baseline covariates
p_w <- 0.5  # probability of a success ("1") in the baseline variables
tx_mult <- 2  # multiplier for the effect of W = 1 on the treatment

## baseline covariate -- simple, binary
W <- as.numeric(replicate(n_w, rbinom(n_obs, 1, p_w)))

## create treatment based on baseline W
A <- as.numeric(rnorm(n_obs, mean = tx_mult * W, sd = 1))

# create outcome as a linear function of A, W + white noise
Y <- A + W + rnorm(n_obs, mean = 0, sd = 1)

# fit the TMLE
tmle_shift <- tmle_medshift(W = W, A = A, Y = Y, delta = 0.5,
                           g_fit_args = list(fit_type = "glm",
                                             nbins = 25,
                                             bin_method = "dhist",
                                             bin_estimator =
                                               speedglmR6$new(),
                                             parfit = FALSE),
                           Q_fit_args = list(fit_type = "glm",
                                             glm_formula = "Y ~ .")
                          )

# conveniently summarize the results
summary(tmle_shift)
```

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/nhejazi/medshift/issues).

---

## Contributions

Contributions are very welcome. Interested contributors should consult our
[contribution
guidelines](https://github.com/nhejazi/medshift/blob/master/CONTRIBUTING.md)
prior to submitting a pull request.

---

## Citation

After using the `medshift` R package, please cite the following:

        @manual{hejazi2018medshift,
          author = {Hejazi, Nima S and {van der Laan}, Mark J and D{\'\i}az,
            Iv{\'a}n},
          title = {medshift: Efficient Estimation of the Causal Mediation
            Effects of Stochastic Interventions in {R}},
          year  = {2018},
          url = {https://github.com/nhejazi/medshift},
          note = {R package version 0.0.1}
        }

---

## License

&copy; 2018 [Nima S. Hejazi](https://nimahejazi.org)

The contents of this repository are distributed under the MIT license. See below
for details:
```
MIT License

Copyright (c) 2018 Nima S. Hejazi

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all
copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
SOFTWARE.
```

---

## References

