---
title: "Causal Mediation Analysis for Stochastic Interventions"
author: "[Nima Hejazi](https://nimahejazi.org) and
  [Iván Díaz](https://www.idiaz.xyz/)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: vignette-refs.bib
vignette: >
  %\VignetteIndexEntry{IPCW-TMLEs with Stochastic Treatment Regimes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r prelim, message=FALSE, warning=FALSE}
# plotting and scientific notation
options(scipen = 999)  ## has scientific notation ever annoyed you?

# packages
library(data.table)
library(dplyr)
library(sl3)
library(medshift)

# prng
set.seed(429153)

# convenience function to compute inference via delta method: EY1 - EY0
param_linear_contrast <- function(params, eifs, ci_level = 0.95) {
  # bounds for confidence interval
  ci_norm_bounds <- c(-1, 1) * abs(stats::qnorm(p = (1 - ci_level) / 2))
  param_est <- params[[1]] - params[[2]]
  eif <- eifs[[1]] - eifs[[2]]
  se_eif <- sqrt(var(eif) / length(eif))
  param_ci <- param_est + ci_norm_bounds * se_eif
  # parameter and inference
  out <- c(param_ci[1], param_est, param_ci[2])
  names(out) <- c("lwr_ci", "param_est", "upr_ci")
  return(out)
}
```

# Background

We are interested in assessing the Natural Direct Effect (NDE) and the Natural
Indirect Effect (NIE), based on the decomposition of the population mediated
intervention mean given in @diaz2019+causal.

To proceed, we'll use as our running example a simple data set from an
observational study of the relationship between BMI and kids behavior,
distributed as part of the [`mma` R package on
CRAN](https://CRAN.R-project.org/package=mma). First, let's load this data set
and take a quick look at it

```{r load_data, message=FALSE, warning=FALSE}
library(mma)
data(weight_behavior)
dim(weight_behavior)
head(weight_behavior)
```

The documentation for the data set describes it as a "database obtained from the
Louisiana State University Health Sciences Center, New Orleans, by  Dr. Richard
Scribner. He  explored the relationship  between BMI and kids behavior through a
survey at children, teachers and parents in Grenada in 2014. This data set
includes 691 observations and 15 variables."

Unfortunately, the data set contains a few observations with missing values. As
these are unrelated to the object of our analysis, we'll simply remove these for
the time being. Note that in a real data analysis, we might consider strategies
to fully make of the observed data, perhaps by imputing missing values. For now,
we simply remove the incomplete observations, resulting in a data set with fewer
observations but much the same structure as the original:

```{r remove_na, echo=FALSE, message=FALSE, warning=FALSE}
# remove missing values
is_na <- unique(do.call(c, apply(apply(weight_behavior, 2, is.na), 2, which)))
weight_behavior_complete <- weight_behavior[-is_na, ]
weight_behavior_complete$sports <-
  as.numeric(weight_behavior_complete$sports) - 1
dim(weight_behavior_complete)
head(weight_behavior_complete)
```

For the analysis of this observational data set, we focus on the effect of
participating in a sports team (`sports`) on the BMI of children (`bmi`), taking
several related covariates as mediators (`snack`, `exercises`, `overweigh`) and
all other collected covariates as potential confounders. Considering an NPSEM,
we separate the observed variables from the data set into their corresponding
nodes as follows

```{r npsem, message=FALSE, warning=FALSE}
Y <- weight_behavior_complete$bmi
A <- weight_behavior_complete$sports
Z <- weight_behavior_complete %>%
  select(snack, exercises, overweigh)
W <- weight_behavior_complete %>%
  select(age, sex, race, numpeople, car, gotosch, tvhours, cmpthours,
         cellhours, sweat)
```

Finally, in our analysis, we consider an incremental propensity score
intervention (IPSI), as first proposed by @kennedy2017nonparametric, wherein the
_odds of participating in a sports team_ is modulated by some fixed amount
($0 \leq \delta \leq 1$) for each individual. Such an intervention may be
interpreted as the effect of a school program that motivates children to
participate in sports teams. To exemplify our approach, we postulate a
motivational intervention that increases the odds of participating in a sports
team by 25% for each individual:

```{r delta_ipsi, message=FALSE, warning=FALSE}
delta_shift_ipsi <- 0.25
```

To easily incorporate ensemble machine learning into the estimation procedure,
we rely on the facilities provided in the [`sl3` R
package](https://tlverse.org/sl3) [@coyle2018sl3]. For a complete guide on using
the `sl3` R package, consider consulting https://tlverse.org/sl3, or
https://tlverse.org (and https://github.com/tlverse) for the `tlverse`
ecosystem, of which `sl3` is a major part. We construct an ensemble learner
using a handful of popular machine learning algorithms below

```{r make_sl, message=FALSE, warning=FALSE}
# random forest learner based on ranger
rf_lrnr_ntree1e3 <- Lrnr_ranger$new(num.trees = 1000)

# SL learners used for continuous data (the nuisance parameter M)
enet_contin_lrnr <- Lrnr_glmnet$new(alpha = 0.5, family = "gaussian",
                                    nfolds = 3)
lasso_contin_lrnr <- Lrnr_glmnet$new(alpha = 1, family = "gaussian",
                                     nfolds = 3)
fglm_contin_lrnr <- Lrnr_glm_fast$new(family = gaussian())
contin_lrnr_lib <- Stack$new(enet_contin_lrnr, lasso_contin_lrnr,
                             fglm_contin_lrnr, rf_lrnr_ntree1e3)
sl_contin_lrnr <- Lrnr_sl$new(learners = contin_lrnr_lib,
                              metalearner = Lrnr_nnls$new())

# SL learners used for binary data (nuisance parameters G and E in this case)
enet_binary_lrnr <- Lrnr_glmnet$new(alpha = 0.5, family = "binomial",
                                    nfolds = 3)
lasso_binary_lrnr <- Lrnr_glmnet$new(alpha = 1, family = "binomial",
                                     nfolds = 3)
fglm_binary_lrnr <- Lrnr_glm_fast$new(family = binomial())
binary_lrnr_lib <- Stack$new(enet_binary_lrnr, lasso_binary_lrnr,
                             fglm_binary_lrnr, rf_lrnr_ntree1e3)
sl_binary_lrnr <- Lrnr_sl$new(learners = binary_lrnr_lib,
                              metalearner = Lrnr_nnls$new())
```

## Decomposing the population intervention mean

We may decompose the PIE in terms of a \textit{direct
effect (DE)} and an \textit{indirect effect (IE)}:
\begin{equation*}
    \psi(\delta) = \overbrace{\mathbb{E}\{Y(g, q) -
                                  Y(g_\delta, q)\}}^{\text{DE}} +
                   \overbrace{\mathbb{E}\{Y(g_\delta, q) -
                                  Y(g_\delta, q_\delta)\}}^{\text{IE}}
\end{equation*}
This decomposition of the PIE as the sum of direct and indirect
effects has an interpretation analogous to the corresponding standard
decomposition of the average treatment effect. In the sequel, we will compute
each of the components of the direct and indirect effects above using
appropriate estimators as follows

* For $\mathbb{E}\{Y(g, q)\}$, the sample mean $\frac{1}{n}\sum_{i=1}^n Y_i$ is
  sufficient;
* for $\mathbb{E}\{Y(g_{\delta}, q)\}$, a one-step efficient estimator for the
  pure effect (of altering the exposure mechanism but not the mediation
  mechanism), as proposed in @diaz2019+causal; and,
* for $\mathbb{E}\{Y(g_{\delta}, q_{\delta})\}$, a one-step efficient estimator
  for the joint effect (of altering both the exposure and mediation mechanisms),
  as proposed in @kennedy2017nonparametric and implemented in the [`npcausal` R
  package](https://github.com/ehkennedy/npcausal).

## Estimating the pure non-mediated effect

As given in @diaz2019+causal, the statistical functional identifying the pure
effect $\mathbb{E}\{Y(g_{\delta}, q)\}$, which corresponds to altering the
exposure mechanism while keeping the mediation mechanism (and its reliance on
the exposure) fixed, is
\begin{equation*}
\theta_0(\delta) = \int m_0(a, z, w) g_{0,\delta}(a \mid w) p_0(z, w)
d\nu(a, z, w),
\end{equation*}
for which a one-step nonparametric-efficient estimator is available. The
corresponding _efficient influence function_ (EIF) with respect to the
nonparametric model $\mathcal{M}$ is $D_{\eta,\delta}(o) = D^Y_{\eta,\delta}(o)
+ D^A_{\eta,\delta}(o) + D^{Z,W}_{\eta,\delta}(o) - \theta(\delta)$. The
one-step estimator may be computed using the EIF estimating equation, making use
of cross-fitting to circumvent any need for entropy conditions. The resultant
estimator is
\begin{equation*}
  \hat{\theta}(\delta) = \frac{1}{n} \sum_{i = 1}^n D_{\hat{\eta}_{j(i)},
  \delta}(O_i) = \frac{1}{n} \sum_{i = 1}^n \left\{ D^Y_{\hat{\eta}_{j(i)},
  \delta}(O_i) + D^A_{\hat{\eta}_{j(i)}, \delta}(O_i) +
  D^{Z,W}_{\hat{\eta}_{j(i)}, \delta}(O_i) \right\},
\end{equation*}
which is implemented in the `medshift` R package [@hejazi2018+medshift]. We
make use of that implementation to estimate $\mathbb{E}\{Y(g_{\delta}, q)\}$
via its one-step estimator $\hat{\theta}(\delta)$ below

```{r efficient_est, message=FALSE, warning=FALSE, cache=TRUE}
# let's compute the parameter where A (but not Z) are shifted
theta_eff <- medshift(W = W, A = A, Z = Z, Y = Y,
                      delta = delta_shift_ipsi,
                      g = sl_binary_lrnr,
                      e = sl_binary_lrnr,
                      m = sl_contin_lrnr,
                      phi = fglm_contin_lrnr,
                      estimator = "onestep",
                      estimator_args = list(cv_folds = 2))
theta_eff
```

## Estimating the Natural Direct Effect

Recall that, based on the decomposition outlined previously, the _natural direct
effect_ (NDE) may be denoted $\beta_{\text{NDE}}(\delta) = \mathbb{E}Y -
\theta_0(\delta)$. Thus, an estimator of the NDE,
$\hat{\beta}_{\text{NDE}}(\delta)$ may be expressed as a composition of
estimators of its constituent parameters:
\begin{equation*}
  \hat{\beta}_{\text{NDE}}({\delta}) = \frac{1}{n} \sum_{i = 1}^n Y_i -
  \hat{\theta}(\delta).
\end{equation*}
Based on the above, we may construct an estimator of the NDE using the
quantities already computed

```{r comp_nde_binary, message=FALSE, warning=FALSE}
# parameter estimates and EIFs for components of direct effect
EY <- mean(Y)
eif_EY <- Y - EY
params_nde <- list(EY, theta_eff$theta)
eifs_nde <- list(eif_EY, theta_eff$eif)

# natural direct effect = EY - estimated quantity
nde_est <- param_linear_contrast(params_nde, eifs_nde)
nde_est
```

As given above, we have for our estimate of the natural direct effect
$\hat{\beta}_{\text{NDE}}({\delta}) =$ `r round(nde_est, 3)`.

## References

